Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2024-08-08 paper/main.tex
Updates

In this section, we develop a novel measure of partial association for mixed
data by extending partial correlations \citep{Whittaker2009} \todo{Verify this
citation} to mixed data. To compute partial correlations we use: 1) A
residualization method, and 2) a measure of associations between the residuals.
In the case of continuous data, the residuals can be computed using a
regression based method and Pearson's correlation coefficient can be used for
computing the association between the residuals. We extend this idea to mixed
data by combining a mixed data residualization method \citep{Ankan2023} with a
multivariate measure of association based, Pillai's Trace \citep{Pillai1955}.
Given $ n $ samples $ D = (x, y, \bm{z}) $ of variables $ X $, $ Y $, and $
\bm{Z} $, we are interested in computing an estimate of the conditional
association $ \rho(X, Y; \bm{Z}) $ between $ X $ and $ Y $ conditioned on $
\bm{Z} $. As the residualization approach requires atleast one conditional
variable to be present, if $ \bm{Z} = \emptyset $, we use a constant value for
$ Z $. Similar to partial correlation, we first compute the residuals for $ X $
and $ Y $ using $ \bm{Z} $ as the predictor variables. We use the mixed data
residualization approach from \citet{Ankan2023}. Depending on the type of
dependent variable, the residuals are defined as follows:

@@

Here, we have the option to choose the estimators based on the properties of
their data. Non-parametric ensemble estimators such as Random Forest and
XGBoost can work on any combination of data types while performing reasonably
well. For linear datasets we can also choose variants of linear regression
depending on the type of data. After the above steps are repeated for both $ X
$ and $ Y $, we end up with two residual matrices: $ R_X $ and $ R_Y $. The
type of the variable determines the shape for these matrices, i.e., if the
variable is continuous we get a residual matrix of shape $ ( n \times 1 ) $ and
if the variable is categorical we have a residual matrix of shape $ ( n \times
(k-1)) $ (where one of the dummy encoded variables is dropped to avoid
multicollinearity)

@@

\begin{definition}
	Given two sets of random variables $ \bm{U} = (U_1, U_2, \cdots, U_p) $
	and $ \bm{V} = (V_1, V_2, \cdots, V_q) $, canonical correlation between
	them, $\phi(\bm{U}, \bm{V}) $ is defined as:

	\begin{equation}
		\phi(\bm{U}, \bm{V}) = \sup_{a, b} \frac{a^T \Sigma_{\bm{U}\bm{V}} b}{\sqrt{a^T \Sigma_{\bm{U}\bm{U}} a} \sqrt{b^T \Sigma_{\bm{V}\bm{V}} b}}
	\end{equation}

@@

>>>>

In this section, we develop a novel measure of partial association for mixed
data by extending partial correlations \citep{Whittaker2009} \todo{Verify this
citation} to mixed data. To compute partial correlations we use: 1) A
residualization method, and 2) a measure of associations between the residuals.
In the case when both $ X $ and $ Y $ are continuous variables, the residuals
are typically computed using a regression based method and  Pearson's
correlation coefficient is used as a measure of association between the
residuals. To extend this idea to mixed data, we combining a mixed data
residualization method \citep{Ankan2023} with a multivariate measure of
association based, Pillai's Trace \citep{Pillai1955}.

Given $ n $ samples $ D = (x, y, \bm{z}) $ of variables $ X $, $ Y $, and $
\bm{Z} $, we are interested in computing an estimate of the conditional
association $ \rho(X, Y; \bm{Z}) $ between $ X $ and $ Y $ conditioned on $
\bm{Z} $. To compute the residuals, we utilize the residualization approach
from \citet{Ankan2023}. Depending on the type of dependent variable, the
residuals are defined as follows:

@@

We repeat the above step for the variable $ Y $ and compute its residual $ R_Y
$. Here, we have the option to choose the estimators based on the properties of
their data. Non-parametric ensemble estimators such as Random Forest and
XGBoost can work on any combination of data types while performing reasonably
well. For linear datasets we can also choose variants of linear regression
depending on the type of data. After the above steps are repeated for both $ X
$ and $ Y $, we end up with two residual matrices: $ R_X $ and $ R_Y $. The
type of the variable determines the shape for these matrices, i.e., if the
variable is continuous we get a residual matrix of shape $ ( n \times 1 ) $ and
if the variable is categorical we have a residual matrix of shape $ ( n \times
(k-1)) $ (where one of the dummy encoded variables is dropped to avoid
multicollinearity)

@@

\begin{definition}

	Given two sets of random variables $ \bm{U} = (U_1, U_2, \cdots, U_p) $
	and $ \bm{V} = (V_1, V_2, \cdots, V_q) $, canonical correlation between
	them, $\phi(\bm{U}, \bm{V}) $ is defined as:

	% \begin{equation}
	% 	\phi(\bm{U}, \bm{V}) = \sup_{a, b} \frac{a^T \Sigma_{\bm{U}\bm{V}} b}{\sqrt{a^T \Sigma_{\bm{U}\bm{U}} a} \sqrt{b^T \Sigma_{\bm{V}\bm{V}} b}}
	% \end{equation}

\begin{equation*}
	\begin{split}
		\bm{\rho}(\bm{U}, \bm{V}) &= \sqrt{\mathrm{E} \left( \Sigma_{\bm{UU}}^{-1} \Sigma_{\bm{UV}} \Sigma_{\bm{VV}}^{-1} \Sigma_{\bm{VU}} \right)}
	\end{split}
\end{equation*}

@@

where $ \mathrm{E}(\Sigma) $ denotes the eigenvalues of the matrix $ \Sigma $.
$ \Sigma_{UU} $ and $ \Sigma_{VV} $ are the covariance matrices of $ \bm{U} $
and $ \bm{V} $, and $ \Sigma_{UV} $ and $ \Sigma_{VU} $ are the
cross-covariance matrices.
